### 2023-02-15 chapter1/chapter1.tex
Removed section numbering from chapter1.tex and reorganized imports

\section*{Introduction}
Comparative genomics is a powerful tool for yielding insights into evolutionary relationships, molecular function, and the forces that drive gene, genome, and population evolution. These methods often rely on the identification of homologous sequences or homologs, that is sequences with common ancestry, since this ensures that differences between sequences reflect variations in evolution from a common point of divergence. However, many analyses impose the additional condition that the sequences have diverged through speciation events (orthology) rather than duplications (paralogy) or other mechanisms such as horizontal gene transfer~\cite{Fitch1970}. The underlying assumption is ``orthologs'' have conserved equivalent functions whereas ``paralogs,'' by virtue of their redundancy, are more likely to diverge~\cite{Ohno1970, Nowak1997, Altenhoff2012, Pegueroles2013, Soria2014}.\footnote{For multiple sequences, orthology is usually defined relative to their most recent common ancestor. This technically includes sequences which split by duplication after this point (``in-paralogs''), but excludes sequences which split by duplication before (``out-paralogs'') \cite{Remm2001}. Many current orthology inference pipelines explicitly incorporate steps to detect in-paralogs. However, the resulting orthologs groups can easily be restricted to those without in-paralogs (``single copy orthologs'') for analyses where an assumption of conserved function is necessary.} This relationship between orthology and function is an essential component of modern biological research since it permits the transfer of annotations between biological systems using sequence similarity alone.

@@

\section*{Results}
\subsection*{Pipeline overview}
Our pipeline follows a similar overall approach to other graph-based methods of orthology inference. First, protein sequences from annotated genomes are collected (Fig.~\ref{fig:graphical_overview}A), and homology searches are conducted between all query-target pairs of genomes (Fig.~\ref{fig:graphical_overview}B). The raw output from these homology searches is processed to yield best hits between pairs of sequences (Fig.~\ref{fig:graphical_overview}C). Next, the network of best hits is clustered into self-consistent orthologous groups (Fig.~\ref{fig:graphical_overview}D). Since genes can have multiplied associated isoforms, we then implemented a novel second clustering step where orthologous groups are grouped by their parent genes, which are represented by the two sets of clusters with warm and cool colors, respectively (Fig.~\ref{fig:graphical_overview}E). Finally, representative sequences in each orthologous group are aligned (Fig.~\ref{fig:graphical_overview}F). In the following sections, we discuss each of these and other steps which were omitted for clarity in greater detail.

\subsection*{Input genomes and pre-processing}
All annotated genomes in the genus \textit{Drosophila} available in April 2022 were downloaded from NCBI's RefSeq database. The assemblies annotated by the NCBI eukaryotic genome annotation pipeline have passed several quality checks and all have supporting transcript evidence, so the annotations are generally highly complete~\cite{Thibaud-Nissen2013}. The \textit{D. miranda} annotation was excluded due to its unusual karyotype~\cite{Dobzhansky1935}. Other annotations were excluded after preliminary clustering showed a deficiency in the number of orthologous groups containing those genomes, indicating their annotations were less complete (data not shown). The \textit{D. melanogaster} annotation was downloaded from FlyBase~\cite{Gramates2022}. In total, the input data consists of 33 genomes, which are listed in Table~\ref{stable:genomes}. Many genes have transcripts that differ only in their UTRs, and as a result there are many duplicate protein sequences in the annotations. Though not strictly necessary, we removed the duplicates in our pipeline, which greatly reduced the computational burden of later steps.

\subsection*{Extraction of best hits from BLAST output}
The protein sequences in each genome annotation were searched against each other in reciprocal pairs using BLAST, yielding a list of high-scoring segment pairs (HSPs) for each query-target pair~\cite{Camacho2009}. HSPs are local alignments, meaning they do not necessarily span the entire lengths of the query and target sequences. Consequently, the search algorithm may return multiple HSPs for each query-target pair if statistically significant regions of homology are separated by nonhomologous or poorly conserved regions. Though the most significant HSP is often used to represent all HSPs between a query-target pair, this approach can fail to rank the pairs by their overall significance if their alignments are broken into multiple HSPs. Furthermore, since query-target pairs were later filtered by the amount overlap between their sequences, it can also exclude pairs that pass the overlap threshold even if the most significant HSP alone does not. Thus, HSPs were merged into a single object called a hit. The best hits for each query were then taken as the highest-scoring hits that passed a minimum overlap criterion and were reciprocal between the query and target sequences.

\subsection*{Clustering in orthologous groups}
The best hits between sequences are naturally visualized as a graph where sequences are nodes and best hits are edges between nodes. Two connected components, sets of nodes joined by a sequence of edges, are shown (Fig.~\ref{fig:components}A,B). The sequences in the first (Fig.~\ref{fig:components}A) all contain C2H2 zinc fingers, whereas the sequences in the second (Fig.~\ref{fig:components}B) are members of the Par-1 family of serine/threonine protein kinases. In both components, some sets of nodes have a high density of edges, forming distinct clusters, whereas other nodes are only sparsely connected to their neighbors. To better understand the structure of these two components, we calculated the number of sequences, unique genes, and unique species in each. The first has 385, 346, and 33 sequences, genes, and species, respectively, and the second has 222, 33, and 33 sequences, genes, and species, respectively. We then plotted the relationship between the number sequences and unique genes across all components to see if this pattern holds true generally (Fig.~\ref{fig:components}C). Two distinct trendlines are apparent. The first increases linearly with the number of sequences with a slope of one, indicating each sequence is generally associated with a unique gene. The second is constant with an intercept of 33, indicating the number of unique genes quickly saturates at the total number of genomes. Thus, there are generally two classes of components. The first is composed of many distinct genes, whereas the second is composed of many different isoforms of a single group of genes.

@@

\subsection*{Addition of paralogs to orthologous groups}
A weakness of the best reciprocal hits criterion is its exclusion of recently diverged paralogs. Since only the highest scoring hits for each query are included in the graph, a paralog without a corresponding duplicate in the target genome is ignored if it is marginally more diverged than the other copy. This is corrected by adding likely paralogs to the orthologous groups. Briefly, the protein sequences in each genome were searched against themselves. If the bit score for an intra-genome hit exceeded the bit score of any inter-genome hits for the same query, the two sequences were identified as a paralogous pair. Orthologous groups were then supplemented with paralogs by adding the paired sequences for each of the original members of the orthologous group. Most orthologous groups contain no paralogs, and those that do generally have few relative to the original number of sequences in the group (Fig.~\ref{sfig:paralogs}).

\subsection*{Grouping orthologous groups by gene}
As genes can have several annotated isoforms, each gene can be associated with several orthologous groups. However, the orthologous groups are not organized into gene-level units since they were clustered using sequence similarity only. A graph-based approach was therefore used to group orthologous groups with similar sets of parent genes. First, a gene overlap graph was constructed by defining an edge between orthologous groups if the intersection of their associated sets of parent genes is at least 50\% of the smaller of the two. Gene groups were then taken as the connected components of the resulting graph, yielding 14,909 groups. This is commensurate with the roughly 15,000 genes in each genome, which suggests this approach has successfully clustered orthologous groups derived from a common set of parent genes.

\subsection*{Initial alignment and selection of representative sequences}
Since the NCBI annotation pipeline incorporates transcriptome data from a variety of sources, its inputs are heterogeneous in sequencing depth, developmental stage, and tissue of origin across different genomes. As a result, some genomes are annotated with different or multiple splice isoforms of a given orthologous gene, which can create complex networks in the resulting hit graph. For example, if the genomes are variably annotated with one or both of two distinct isoforms, the resulting graph may contain two clusters connected by a ``bridge'' formed by the genomes which contain only one of the isoforms. If the nodes bridging the two clusters form cliques with themselves and the clusters, the clique percolation algorithm will merge all the nodes into a single orthologous group where some genes have multiple associated sequences. However, these additional sequences can complicate downstream comparative analyses that may not easily generalize to genes with multiple associated sequences. Thus, in our pipeline a single representative was chosen for each gene using an alignment-based strategy detailed in the methods section. Briefly, a statistical profile was created from an alignment of the sequences in each orthologous group, and the representative for each gene was chosen as the sequence which best matched this profile.

\subsection*{Selection of single copy orthologous groups}
The criteria for selecting orthologous groups for further analyses depends on the biological question under investigation. For example, studies of gene duplication will focus on orthologous groups with paralogs in some lineages but not in others. In contrast, analyses which assume functional conservation should restrict the orthologous groups to single copy orthologs since paralogs more frequently undergo functional divergence~\cite{Altenhoff2012, Pegueroles2013, Soria2014}. A simple method for identifying such groups is requiring each species to have exactly one associated gene. However, since the probability of at least one missing gene annotation approaches one as the total number of genomes increases, this is too restrictive and fails to leverage the redundancy of closely related genomes. Instead, a set of phylogenetic diversity criteria detailed in Table~\ref{stable:diversity_criteria} were applied to ensure the major lineages were represented in downstream analyses. Furthermore, genome-wide analyses should select one orthologous group per each of the previously identified gene groups as to not bias the results towards genes with many distinct groups of isoforms. In summary, orthologous groups failing the phylogenetic diversity criteria were first removed, and the representative for each gene group was chosen as the highest scoring orthologous group when ranked by the number species and the sum of the bit scores associated with each edge. This significantly reduced the number of orthologous groups from 22,813 to 8,566.

\subsection*{Alignment refinement}
Though the pipeline’s quality control measures ensure a high degree of overall sequence identity between members of an orthologous group, some sequences contain long ``poorly supported'' segments which have no homology to most or any other sequences in the alignment. Since most common multiple sequence alignment algorithms assume the sequences are largely homologous, these segments are sometimes ``over-aligned'' by forcing them into alignment where chance sequence similarities occur. Typically, these segments remain contiguous, so the alignments alternate between short runs of columns with few or no gaps and large gap-rich regions (Fig.~\ref{fig:realignment}A-B, left). More rarely, when long poorly supported segments are adjacent to a long gap in the same sequence, the two are interlaced, yielding long gaps interrupted by short segments of spurious alignment (Fig.~\ref{fig:realignment}C, left).

@@

\subsection*{Alignment curation}
Although the refinement process corrects most cases of over-alignment, the alignment may still contain regions whose aligned segments have poor or inconsistent support. For example, long poorly supported segments in internal regions were not removed from the alignment since they are bounded by at least one consensus column to the left and right. Additionally, some regions have a significant fraction of sequences with strongly supported segments, but the observed gap pattern is discordant with the expected phylogenetic relationships. Since they are present in so few sequences, the former segments are likely artifactual, resulting from errors during assembly or annotation. (Biological explanations such as alternative splice sites, frameshift mutations, or transposition events are also possible, however.) In contrast, the high sequence identity and clear boundaries of the segments in the latter regions suggest they are conserved but skipped exons. Given the heterogeneous sourcing of the transcript evidence, these sequences containing these segments are likely splice isoforms specific to certain tissues or developmental stage.

@@

\subsection*{Inference of species trees}
Many phylogenetic methods require a species tree to inform the evolutionary relationships between sequences. In fact, the phylo-HMMs discussed previously used a species tree as an input, though we omitted this detail for clarity of exposition. Therefore, to support the curation step and other downstream analyses, we sought to infer phylogenetic trees from the aligned sequences. However, since the roots of phylogenetic trees are not identifiable with commonly used time-reversible substitution models, we repeated the orthology inference pipeline with the outgroup species \textit{Scaptodrosophila lebanonensis}. Afterwards, we inferred phylogenetic trees using the LG model of amino acid substitution from 100 meta-alignments sampled from alignments of single copy orthologous groups. We then combined them into a single consensus tree (Fig.~\ref{fig:trees}A). To provide a similar tree for the analysis of non-coding regions, we inferred phylogenetic trees using the GTR model of nucleotide substitution from 100 meta-alignments sampled from nucleotide alignments which were ``reverse translated'' from the protein alignments and their corresponding coding sequences (\ref{fig:trees}B). Both trees have an identical topology, which is consistent with other published phylogenies~\cite{D12GC2007, DaLage2007}.

@@

\section*{Discussion}
The orthologous groups and alignments yielded by this pipeline are a valuable resource for comparative studies of gene birth/death processes and protein evolution at the level of both entire proteomes and specific gene families in the \textit{Drosophila} genus. To assist these efforts, the final alignments after curation and the labels from the ``missing'' phylo-HMM are provided as supplemental data. Although this work focused on single copy orthologs, other studies may require different subsets of orthologous groups that demand other pre-processing and alignment strategies. Therefore, we have included the orthologous groups and the initial alignments with and without non-representative sequences in the supplemental data as well. While we anticipate these resources will remain relevant in the near term, the trends that permitted this work to substantially improve on previous efforts will render them obsolete in the coming years as more \textit{Drosophila} genomes are assembled and annotated. However, an authoritative and lasting set of orthologous groups and alignments is not the primary goal of this work. Instead, it serves as a case study in how dense taxonomic sampling and modern genome assembly and annotation pipelines present new opportunities and challenges to the traditional techniques for identifying and aligning orthologous groups.

@@

\section*{Materials and methods}
\subsection*{Sequence de-duplication and BLAST search parameters}
The protein sequences for each annotation were de-duplicated by removing any sequences which had already appeared in association with the same gene. Thus, the first accession associated with a sequence and gene pair was the sequence’s representative accession for the gene. BLAST+ 2.13.0 was used for the sequence similarity searches~\cite{Camacho2009}. An E-value cutoff of 1 was used for the initial searches. However, this cutoff was lowered to 1E-10 during processing of the BLAST output.

\subsection*{Extraction of HSPs from BLAST output}
To reduce the computational burden of merging HSPs into hits, the BLAST output was filtered to extract HSPs associated with the highest-scoring gene. The HSPs were first grouped by target protein, and the resulting groups were sorted in descending order by the bit score of their highest-scoring HSP. Iterating over the groups, all HSPs in a group were passed to the next step until the parent gene of the group was not the parent gene of the highest ranked group. This method collected all candidate HSPs for a target gene if the highest-scoring HSP within a group exceeded the highest-scoring HSP of the next best gene. This is in some senses an extension of the best hit criterion where hits are considered at the level of genes rather than proteins. If multiple genes tied for the highest-scoring HSP, the iteration stopped when the parent gene of the current group matched none of these highest-scoring genes.

\subsection*{Merging of HSPs into hits}
HSPs were merged in two stages where the first combined non-overlapping HSPs, and the second combined the remaining HSPs. In the first stage, proceeding from highest to lowest bit score, HSPs were marked as ``disjoint'' if they did not overlap with any other HSP previously marked as disjoint. Although this greedy strategy did not necessarily yield the highest-scoring set of disjoint HSPs, it prioritized higher scoring HSPs. In the second stage, all disjoint HSPs were marked as ``compatible,'' and proceeding from highest to lowest bit score the remaining HSPs were marked as compatible if the overlap with any other compatible HSP was no more than 50\% of the length of either. The best hit for each query was chosen as the hit with the highest sum of bit scores from disjoint HSPs. The best hits were filtered by overlap and reciprocity criteria. The overlap criterion was applied first and required that 50\% of residues in the query were aligned in compatible HSPs. This excluded false positives from conserved domains embedded in larger non-homologous proteins by ensuring the hits spanned a sufficient fraction of the query and target sequences. The reciprocity criterion required each query-target pair had a corresponding hit where the roles were reversed, which ensured there was no ambiguity in which target was the best match for the query.

\subsection*{Clustering by \textit{k}-clique percolation}
\textit{k}-clique percolation was implemented in two steps. In the first, maximal cliques were identified. In the second, a percolation graph was constructed by defining edges between cliques if they had \textit{k}-1 nodes in common. Clusters were the connected components of this second graph. The first step used the NetworkX implementation of a maximal clique algorithm. The second step used a modification of the NetworkX implementation of the \textit{k}-clique community algorithm. The NetworkX implementation exhaustively finds all edges in the percolation graph. Since joining a \textit{k}-clique community only requires that a clique has a single edge connecting it to that a community, this approach was needlessly expensive for large graphs. The custom implementation instead used a progressive approach where each clique was checked against a list of known communities, merging communities as necessary in each step.

@@

\subsection*{Addition of paralogs to orthologous groups}
The protein sequences for each annotation were searched against themselves with the same settings as for the inter-genome searches. The resulting output was processed identically except the HSPs were not filtered using the best gene criterion. Thus, all HSPs for each query were merged into hits. The best hit for each query and target gene was chosen as the hit with the highest sum of bit scores from disjoint HSPs. (Grouping by target gene ensured only the highest-scoring isoform was selected.) Query-target pairs whose hits exceeded the maximum bit score for all inter-genome hits associated with that query and passed the overlap and reciprocity filters were designated as paralogous pairs. The orthologous groups were supplemented with paralogs by adding the paired sequences for each of the original members of the orthologous group.

\subsection*{Initial alignment and selection of representative sequences}
The sequences in each orthologous group were aligned using MAFFT 7.490 with the following settings: -{}-globalpair -{}-maxiterate 1000 -{}-thread 1 -{}-anysymbol -{}-allowshift -{}-leavegappyregion -{}-unalignlevel 0.4~\cite{Katoh2013}. Representative sequences for each gene were selected by maximum likelihood according to binary profiles constructed from these alignments. First each sequence was coded into gap and non-gap symbols. The sequences were grouped by gene, and for each group and position if at least one sequence was aligned in the group, the group contributed one count for the non-gap symbol to the profile at that position. Otherwise, the group contributed a count for the gap symbol at that position. To account for the phylogenetic dependencies between sequences, the counts were weighted according to a Gaussian process over the GTR2 consensus tree described in the section on inferring species trees~\cite{Altschul1989}. If a species had multiple genes in the orthologous group, the species weight was divided evenly among them. Each coded sequence was scored according to this profile, and the maximum likelihood sequence for each gene was selected as its representative. By assigning a non-gap count to groups and positions where at least one sequence was aligned, the profile prioritized the selection of sequences with the fewest gaps that best matched the consensus alignment. Since this scheme can cause a sequence to score negative infinity if it has a gap at a position where every group has at least one aligned sequence, the profile was initialized with a pseudocount of 0.005 for the gap and non-gap symbols at each position.

>>>>

\section{Introduction}
Comparative genomics is a powerful tool for yielding insights into evolutionary relationships, molecular function, and the forces that drive gene, genome, and population evolution. These methods often rely on the identification of homologous sequences or homologs, that is sequences with common ancestry, since this ensures that differences between sequences reflect variations in evolution from a common point of divergence. However, many analyses impose the additional condition that the sequences have diverged through speciation events (orthology) rather than duplications (paralogy) or other mechanisms such as horizontal gene transfer~\cite{Fitch1970}. The underlying assumption is ``orthologs'' have conserved equivalent functions whereas ``paralogs,'' by virtue of their redundancy, are more likely to diverge~\cite{Ohno1970, Nowak1997, Altenhoff2012, Pegueroles2013, Soria2014}.\footnote{For multiple sequences, orthology is usually defined relative to their most recent common ancestor. This technically includes sequences which split by duplication after this point (``in-paralogs''), but excludes sequences which split by duplication before (``out-paralogs'') \cite{Remm2001}. Many current orthology inference pipelines explicitly incorporate steps to detect in-paralogs. However, the resulting orthologs groups can easily be restricted to those without in-paralogs (``single copy orthologs'') for analyses where an assumption of conserved function is necessary.} This relationship between orthology and function is an essential component of modern biological research since it permits the transfer of annotations between biological systems using sequence similarity alone.

@@

\section{Results}
\subsection{Pipeline overview}
Our pipeline follows a similar overall approach to other graph-based methods of orthology inference. First, protein sequences from annotated genomes are collected (Fig.~\ref{fig:graphical_overview}A), and homology searches are conducted between all query-target pairs of genomes (Fig.~\ref{fig:graphical_overview}B). The raw output from these homology searches is processed to yield best hits between pairs of sequences (Fig.~\ref{fig:graphical_overview}C). Next, the network of best hits is clustered into self-consistent orthologous groups (Fig.~\ref{fig:graphical_overview}D). Since genes can have multiplied associated isoforms, we then implemented a novel second clustering step where orthologous groups are grouped by their parent genes, which are represented by the two sets of clusters with warm and cool colors, respectively (Fig.~\ref{fig:graphical_overview}E). Finally, representative sequences in each orthologous group are aligned (Fig.~\ref{fig:graphical_overview}F). In the following sections, we discuss each of these and other steps which were omitted for clarity in greater detail.

\subsection{Input genomes and pre-processing}
All annotated genomes in the genus \textit{Drosophila} available in April 2022 were downloaded from NCBI's RefSeq database. The assemblies annotated by the NCBI eukaryotic genome annotation pipeline have passed several quality checks and all have supporting transcript evidence, so the annotations are generally highly complete~\cite{Thibaud-Nissen2013}. The \textit{D. miranda} annotation was excluded due to its unusual karyotype~\cite{Dobzhansky1935}. Other annotations were excluded after preliminary clustering showed a deficiency in the number of orthologous groups containing those genomes, indicating their annotations were less complete (data not shown). The \textit{D. melanogaster} annotation was downloaded from FlyBase~\cite{Gramates2022}. In total, the input data consists of 33 genomes, which are listed in Table~\ref{stable:genomes}. Many genes have transcripts that differ only in their UTRs, and as a result there are many duplicate protein sequences in the annotations. Though not strictly necessary, we removed the duplicates in our pipeline, which greatly reduced the computational burden of later steps.

\subsection{Extraction of best hits from BLAST output}
The protein sequences in each genome annotation were searched against each other in reciprocal pairs using BLAST, yielding a list of high-scoring segment pairs (HSPs) for each query-target pair~\cite{Camacho2009}. HSPs are local alignments, meaning they do not necessarily span the entire lengths of the query and target sequences. Consequently, the search algorithm may return multiple HSPs for each query-target pair if statistically significant regions of homology are separated by nonhomologous or poorly conserved regions. Though the most significant HSP is often used to represent all HSPs between a query-target pair, this approach can fail to rank the pairs by their overall significance if their alignments are broken into multiple HSPs. Furthermore, since query-target pairs were later filtered by the amount overlap between their sequences, it can also exclude pairs that pass the overlap threshold even if the most significant HSP alone does not. Thus, HSPs were merged into a single object called a hit. The best hits for each query were then taken as the highest-scoring hits that passed a minimum overlap criterion and were reciprocal between the query and target sequences.

\subsection{Clustering in orthologous groups}
The best hits between sequences are naturally visualized as a graph where sequences are nodes and best hits are edges between nodes. Two connected components, sets of nodes joined by a sequence of edges, are shown (Fig.~\ref{fig:components}A,B). The sequences in the first (Fig.~\ref{fig:components}A) all contain C2H2 zinc fingers, whereas the sequences in the second (Fig.~\ref{fig:components}B) are members of the Par-1 family of serine/threonine protein kinases. In both components, some sets of nodes have a high density of edges, forming distinct clusters, whereas other nodes are only sparsely connected to their neighbors. To better understand the structure of these two components, we calculated the number of sequences, unique genes, and unique species in each. The first has 385, 346, and 33 sequences, genes, and species, respectively, and the second has 222, 33, and 33 sequences, genes, and species, respectively. We then plotted the relationship between the number sequences and unique genes across all components to see if this pattern holds true generally (Fig.~\ref{fig:components}C). Two distinct trendlines are apparent. The first increases linearly with the number of sequences with a slope of one, indicating each sequence is generally associated with a unique gene. The second is constant with an intercept of 33, indicating the number of unique genes quickly saturates at the total number of genomes. Thus, there are generally two classes of components. The first is composed of many distinct genes, whereas the second is composed of many different isoforms of a single group of genes.

@@

\subsection{Addition of paralogs to orthologous groups}
A weakness of the best reciprocal hits criterion is its exclusion of recently diverged paralogs. Since only the highest scoring hits for each query are included in the graph, a paralog without a corresponding duplicate in the target genome is ignored if it is marginally more diverged than the other copy. This is corrected by adding likely paralogs to the orthologous groups. Briefly, the protein sequences in each genome were searched against themselves. If the bit score for an intra-genome hit exceeded the bit score of any inter-genome hits for the same query, the two sequences were identified as a paralogous pair. Orthologous groups were then supplemented with paralogs by adding the paired sequences for each of the original members of the orthologous group. Most orthologous groups contain no paralogs, and those that do generally have few relative to the original number of sequences in the group (Fig.~\ref{sfig:paralogs}).

\subsection{Grouping orthologous groups by gene}
As genes can have several annotated isoforms, each gene can be associated with several orthologous groups. However, the orthologous groups are not organized into gene-level units since they were clustered using sequence similarity only. A graph-based approach was therefore used to group orthologous groups with similar sets of parent genes. First, a gene overlap graph was constructed by defining an edge between orthologous groups if the intersection of their associated sets of parent genes is at least 50\% of the smaller of the two. Gene groups were then taken as the connected components of the resulting graph, yielding 14,909 groups. This is commensurate with the roughly 15,000 genes in each genome, which suggests this approach has successfully clustered orthologous groups derived from a common set of parent genes.

\subsection{Initial alignment and selection of representative sequences}
Since the NCBI annotation pipeline incorporates transcriptome data from a variety of sources, its inputs are heterogeneous in sequencing depth, developmental stage, and tissue of origin across different genomes. As a result, some genomes are annotated with different or multiple splice isoforms of a given orthologous gene, which can create complex networks in the resulting hit graph. For example, if the genomes are variably annotated with one or both of two distinct isoforms, the resulting graph may contain two clusters connected by a ``bridge'' formed by the genomes which contain only one of the isoforms. If the nodes bridging the two clusters form cliques with themselves and the clusters, the clique percolation algorithm will merge all the nodes into a single orthologous group where some genes have multiple associated sequences. However, these additional sequences can complicate downstream comparative analyses that may not easily generalize to genes with multiple associated sequences. Thus, in our pipeline a single representative was chosen for each gene using an alignment-based strategy detailed in the methods section. Briefly, a statistical profile was created from an alignment of the sequences in each orthologous group, and the representative for each gene was chosen as the sequence which best matched this profile.

\subsection{Selection of single copy orthologous groups}
The criteria for selecting orthologous groups for further analyses depends on the biological question under investigation. For example, studies of gene duplication will focus on orthologous groups with paralogs in some lineages but not in others. In contrast, analyses which assume functional conservation should restrict the orthologous groups to single copy orthologs since paralogs more frequently undergo functional divergence~\cite{Altenhoff2012, Pegueroles2013, Soria2014}. A simple method for identifying such groups is requiring each species to have exactly one associated gene. However, since the probability of at least one missing gene annotation approaches one as the total number of genomes increases, this is too restrictive and fails to leverage the redundancy of closely related genomes. Instead, a set of phylogenetic diversity criteria detailed in Table~\ref{stable:diversity_criteria} were applied to ensure the major lineages were represented in downstream analyses. Furthermore, genome-wide analyses should select one orthologous group per each of the previously identified gene groups as to not bias the results towards genes with many distinct groups of isoforms. In summary, orthologous groups failing the phylogenetic diversity criteria were first removed, and the representative for each gene group was chosen as the highest scoring orthologous group when ranked by the number species and the sum of the bit scores associated with each edge. This significantly reduced the number of orthologous groups from 22,813 to 8,566.

\subsection{Alignment refinement}
Though the pipeline’s quality control measures ensure a high degree of overall sequence identity between members of an orthologous group, some sequences contain long ``poorly supported'' segments which have no homology to most or any other sequences in the alignment. Since most common multiple sequence alignment algorithms assume the sequences are largely homologous, these segments are sometimes ``over-aligned'' by forcing them into alignment where chance sequence similarities occur. Typically, these segments remain contiguous, so the alignments alternate between short runs of columns with few or no gaps and large gap-rich regions (Fig.~\ref{fig:realignment}A-B, left). More rarely, when long poorly supported segments are adjacent to a long gap in the same sequence, the two are interlaced, yielding long gaps interrupted by short segments of spurious alignment (Fig.~\ref{fig:realignment}C, left).

@@

\subsection{Alignment curation}
Although the refinement process corrects most cases of over-alignment, the alignment may still contain regions whose aligned segments have poor or inconsistent support. For example, long poorly supported segments in internal regions were not removed from the alignment since they are bounded by at least one consensus column to the left and right. Additionally, some regions have a significant fraction of sequences with strongly supported segments, but the observed gap pattern is discordant with the expected phylogenetic relationships. Since they are present in so few sequences, the former segments are likely artifactual, resulting from errors during assembly or annotation. (Biological explanations such as alternative splice sites, frameshift mutations, or transposition events are also possible, however.) In contrast, the high sequence identity and clear boundaries of the segments in the latter regions suggest they are conserved but skipped exons. Given the heterogeneous sourcing of the transcript evidence, these sequences containing these segments are likely splice isoforms specific to certain tissues or developmental stage.

@@

\subsection{Inference of species trees}
Many phylogenetic methods require a species tree to inform the evolutionary relationships between sequences. In fact, the phylo-HMMs discussed previously used a species tree as an input, though we omitted this detail for clarity of exposition. Therefore, to support the curation step and other downstream analyses, we sought to infer phylogenetic trees from the aligned sequences. However, since the roots of phylogenetic trees are not identifiable with commonly used time-reversible substitution models, we repeated the orthology inference pipeline with the outgroup species \textit{Scaptodrosophila lebanonensis}. Afterwards, we inferred phylogenetic trees using the LG model of amino acid substitution from 100 meta-alignments sampled from alignments of single copy orthologous groups. We then combined them into a single consensus tree (Fig.~\ref{fig:trees}A). To provide a similar tree for the analysis of non-coding regions, we inferred phylogenetic trees using the GTR model of nucleotide substitution from 100 meta-alignments sampled from nucleotide alignments which were ``reverse translated'' from the protein alignments and their corresponding coding sequences (\ref{fig:trees}B). Both trees have an identical topology, which is consistent with other published phylogenies~\cite{D12GC2007, DaLage2007}.

@@

\section{Discussion}
The orthologous groups and alignments yielded by this pipeline are a valuable resource for comparative studies of gene birth/death processes and protein evolution at the level of both entire proteomes and specific gene families in the \textit{Drosophila} genus. To assist these efforts, the final alignments after curation and the labels from the ``missing'' phylo-HMM are provided as supplemental data. Although this work focused on single copy orthologs, other studies may require different subsets of orthologous groups that demand other pre-processing and alignment strategies. Therefore, we have included the orthologous groups and the initial alignments with and without non-representative sequences in the supplemental data as well. While we anticipate these resources will remain relevant in the near term, the trends that permitted this work to substantially improve on previous efforts will render them obsolete in the coming years as more \textit{Drosophila} genomes are assembled and annotated. However, an authoritative and lasting set of orthologous groups and alignments is not the primary goal of this work. Instead, it serves as a case study in how dense taxonomic sampling and modern genome assembly and annotation pipelines present new opportunities and challenges to the traditional techniques for identifying and aligning orthologous groups.

@@

\section{Materials and methods}
\subsection{Sequence de-duplication and BLAST search parameters}
The protein sequences for each annotation were de-duplicated by removing any sequences which had already appeared in association with the same gene. Thus, the first accession associated with a sequence and gene pair was the sequence’s representative accession for the gene. BLAST+ 2.13.0 was used for the sequence similarity searches~\cite{Camacho2009}. An E-value cutoff of 1 was used for the initial searches. However, this cutoff was lowered to 1E-10 during processing of the BLAST output.

\subsection{Extraction of HSPs from BLAST output}
To reduce the computational burden of merging HSPs into hits, the BLAST output was filtered to extract HSPs associated with the highest-scoring gene. The HSPs were first grouped by target protein, and the resulting groups were sorted in descending order by the bit score of their highest-scoring HSP. Iterating over the groups, all HSPs in a group were passed to the next step until the parent gene of the group was not the parent gene of the highest ranked group. This method collected all candidate HSPs for a target gene if the highest-scoring HSP within a group exceeded the highest-scoring HSP of the next best gene. This is in some senses an extension of the best hit criterion where hits are considered at the level of genes rather than proteins. If multiple genes tied for the highest-scoring HSP, the iteration stopped when the parent gene of the current group matched none of these highest-scoring genes.

\subsection{Merging of HSPs into hits}
HSPs were merged in two stages where the first combined non-overlapping HSPs, and the second combined the remaining HSPs. In the first stage, proceeding from highest to lowest bit score, HSPs were marked as ``disjoint'' if they did not overlap with any other HSP previously marked as disjoint. Although this greedy strategy did not necessarily yield the highest-scoring set of disjoint HSPs, it prioritized higher scoring HSPs. In the second stage, all disjoint HSPs were marked as ``compatible,'' and proceeding from highest to lowest bit score the remaining HSPs were marked as compatible if the overlap with any other compatible HSP was no more than 50\% of the length of either. The best hit for each query was chosen as the hit with the highest sum of bit scores from disjoint HSPs. The best hits were filtered by overlap and reciprocity criteria. The overlap criterion was applied first and required that 50\% of residues in the query were aligned in compatible HSPs. This excluded false positives from conserved domains embedded in larger non-homologous proteins by ensuring the hits spanned a sufficient fraction of the query and target sequences. The reciprocity criterion required each query-target pair had a corresponding hit where the roles were reversed, which ensured there was no ambiguity in which target was the best match for the query.

\subsection{Clustering by \textit{k}-clique percolation}
\textit{k}-clique percolation was implemented in two steps. In the first, maximal cliques were identified. In the second, a percolation graph was constructed by defining edges between cliques if they had \textit{k}-1 nodes in common. Clusters were the connected components of this second graph. The first step used the NetworkX implementation of a maximal clique algorithm. The second step used a modification of the NetworkX implementation of the \textit{k}-clique community algorithm. The NetworkX implementation exhaustively finds all edges in the percolation graph. Since joining a \textit{k}-clique community only requires that a clique has a single edge connecting it to that a community, this approach was needlessly expensive for large graphs. The custom implementation instead used a progressive approach where each clique was checked against a list of known communities, merging communities as necessary in each step.

@@

\subsection{Addition of paralogs to orthologous groups}
The protein sequences for each annotation were searched against themselves with the same settings as for the inter-genome searches. The resulting output was processed identically except the HSPs were not filtered using the best gene criterion. Thus, all HSPs for each query were merged into hits. The best hit for each query and target gene was chosen as the hit with the highest sum of bit scores from disjoint HSPs. (Grouping by target gene ensured only the highest-scoring isoform was selected.) Query-target pairs whose hits exceeded the maximum bit score for all inter-genome hits associated with that query and passed the overlap and reciprocity filters were designated as paralogous pairs. The orthologous groups were supplemented with paralogs by adding the paired sequences for each of the original members of the orthologous group.

\subsection{Initial alignment and selection of representative sequences}
The sequences in each orthologous group were aligned using MAFFT 7.490 with the following settings: -{}-globalpair -{}-maxiterate 1000 -{}-thread 1 -{}-anysymbol -{}-allowshift -{}-leavegappyregion -{}-unalignlevel 0.4~\cite{Katoh2013}. Representative sequences for each gene were selected by maximum likelihood according to binary profiles constructed from these alignments. First each sequence was coded into gap and non-gap symbols. The sequences were grouped by gene, and for each group and position if at least one sequence was aligned in the group, the group contributed one count for the non-gap symbol to the profile at that position. Otherwise, the group contributed a count for the gap symbol at that position. To account for the phylogenetic dependencies between sequences, the counts were weighted according to a Gaussian process over the GTR2 consensus tree described in the section on inferring species trees~\cite{Altschul1989}. If a species had multiple genes in the orthologous group, the species weight was divided evenly among them. Each coded sequence was scored according to this profile, and the maximum likelihood sequence for each gene was selected as its representative. By assigning a non-gap count to groups and positions where at least one sequence was aligned, the profile prioritized the selection of sequences with the fewest gaps that best matched the consensus alignment. Since this scheme can cause a sequence to score negative infinity if it has a gap at a position where every group has at least one aligned sequence, the profile was initialized with a pseudocount of 0.005 for the gap and non-gap symbols at each position.
